---
title: "The ESSP Package"
subtitle: "A plain-English pipeline for U.S. energy data"
author: "Energy Security Studies Program"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    number-sections: true
    code-fold: false
    code-tools: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
    # Match showtext's registration dpi (register_merriweather sets dpi=300),
    # or the Merriweather glyphs render at the wrong physical size.
    fig-dpi: 300
execute:
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false
library(ESSP)
library(knitr)
options(digits = 4)
```

# What this is

`ESSP` is an in-house R package for energy research. It wraps the U.S. Energy Information Administration (EIA) data API behind a small set of plain-English commands, so a student can go from a question to a finished, house-styled chart without touching a URL, an API key, or a raw data file.

Everything below is **live**: each table and figure on this page was pulled straight from EIA when this document was rendered. Nothing is pasted in.

The package is organized around three ideas:

1.  **Get** — `essp.get()`, downloadsmost questions.
2.  **Analyze** — `analyze.*` commands return tidy, analysis-ready tables.
3.  **Chart** — `chart.*` commands draw the program's signature figures.

A student never has to know whether a topic is keyed by state or by grid operator, or how EIA formats its dates — the commands take ordinary strings.

# Downloading Raw Data: `essp.get()`

`essp.catalog()` lists every topic that you can "get", and whether it expects a **state**, a **balancing authority** (grid operator), or a **fuel**.

```{r}
#| label: catalog
# One call lists the whole vocabulary -- no manual to read.
kable(essp.catalog(), caption = "Everything essp.get() can fetch")
```

With that, a single line answers a question. `what` decides how `where` is read, so the student doesn't have to:

```{r}
#| label: get-demo
# "Give me Georgia's generation mix for 2024."
essp.get("mix", "GA", "2024") |> head(8) |>
  kable(digits = 1, caption = "essp.get('mix', 'GA', '2024')")
```

# Analysis tables

The `analyze.*` commands are the engine behind `essp.get()`. Each takes the same simple `(state, year)` form, fills in sensible defaults, and returns a tidy table you can read, export, or feed to a chart.

## Generation vs. capacity mix

This is the base generation and capacity analysis command.

```{r}
#| label: mix-table
# The national picture for 2024. `solar = "total"` counts small rooftop solar.
mix_us <- analyze.mix("US", "2024")
kable(mix_us, digits = 1,
      caption = "U.S. capacity and generation by resource, 2024")
```

## Fuel diversity across states, in one call

Because every metric now accepts **several states at once**, a comparison that used to need a loop is a single line. Here we score three states on how concentrated their power mix is.

I am going to add various calculation functions such as the **HHI** (Herfindahl–Hirschman Index), as we used for wallet-share and other insight reports. The HHI is the standard measure of concentration: add up the square of each fuel's percentage share of the mix. It runs from near 0 (many fuels, evenly balanced) to 10,000 (a single fuel supplies everything). A higher HHI means more dependence on one fuel.

```{r}
#| label: diversity-multi
# Three states, stacked into one table with a `state` column.
analyze.diversity("GA,WV,CA", "2024") |>
  kable(digits = 1,
        caption = "Fuel-mix concentration: GA vs. WV vs. CA, 2024")
```

Use case: West Virginia, running almost entirely on coal, scores as the most concentrated; California, with a broad renewable mix, the least.

## Use case: Who's the leading producer of X?

`analyze.fuelleaders()` answers "which states produce the most of X?" with the cumulative share carried alongside, so a claim like *"the top 8 states account for N% of U.S. solar"* writes itself from the table.

```{r}
#| label: fuelleaders
analyze.fuelleaders("Solar", "2024", n = 8) |>
  kable(digits = 1, caption = "Top solar-generating states, 2024")
```

# Charts

The `chart.*` commands render the program's house figures. They accept the same plain strings — fetch and draw happen in one call.

## Capacity vs. generation

`chart.generation()` is the signature 0–100% stacked bar. Passing a resource name highlights it and greys the rest, keeping the house colors consistent across every figure.

```{r}
#| label: chart-generation
#| fig-cap: "U.S. fleet makeup, solar highlighted (chart.generation)."
chart.generation("Solar", "US", "2024")
```

## The daily demand curve

`chart.demandcurve()` builds the stacked daily generation profile for a grid operator — here California (CISO), averaged across a summer day. It runs the whole pipeline (hourly fuel mix, the natural-gas split into baseload and peakers, and battery storage) behind a single call.

```{r}
#| label: chart-demand
#| fig-cap: "CISO summer demand curve, Solar and NGCT highlighted (chart.demandcurve)."
#| fig-width: 11
#| fig-height: 6
chart.demandcurve("Solar,NGCT", "CISO", "2024")
```

::: callout-note
## A note on batteries in the live data

EIA's hourly grid-monitor feed does not report battery storage as a fuel type, so a curve built strictly from live data shows no storage band, the real `BAT (Battery)` series is unavailable for every region we pull.\
\
The reference script that inspired this chart drew a storage band, but it did so from a *modeled* (synthetic) storage profile rather than measured hourly output.

The original R script imputes a cosine wave:\
\
bump \<- function(h, center, halfwidth, peak) { out \<- numeric(length(h)) win \<- abs(h - center) \< halfwidth out\[win\] \<- peak \* 0.5 \* (1 + cos(pi \* (h\[win\] - center) / halfwidth)) out }\
\
Should I impute this as well and make a note that battery storage isn't lined to data?
:::

# Normative Time and State

**Normative time.** The hourly commands take ordinary time phrases, so nobody types an ISO timestamp:

```{r}
#| label: normative-time
# "Give me CISO's average demand profile for July 2024."
analyze.loadshape("CISO", "July 2024") |> head(6) |>
  kable(digits = 0, caption = "analyze.loadshape('CISO', 'July 2024')")
```

The same command scales up without changing shape — ask for several states and years and it stacks them, keeping `state` and `year` columns so the result stays self-describing:

```{r}
#| label: multi
analyze.capacityfactor("GA,OH", "2023,2024") |>
  kable(digits = 1,
        caption = "Capacity factor by resource: 2 states x 2 years")
```

# In short

A student loads the package and asks questions in plain English:

``` r
library(ESSP)
essp.get("mix", "GA", "2024")          # a table
chart.generation("Solar", "US", "2024") # a figure
chart.demandcurve("Solar", "CISO", "2024")
```

Every number and figure above came from EIA at render time, through those same commands — reproducible, citable, and in the program's house style.

# Future Ideas:

::: {.callout-tip appearance="minimal"}
## 1. an essp.print() command that outputs the code for various figures and tables and plots, in case someone wants access to the backend

2.  essp.write() command that writes excel files and csvs
3.  Plain-language explinations for errors, when stuff is downloading, possible inline embedded AI assistant -- new package tool i found
4.  Quarto-playground similar to canva that allows students to make reports without needing to actually code anything.
5.  Publish to C-RAN eventually under MIT open access license.
:::